Composable data visualizations

### 2021-03-22 wip-compost/paper.tex
Design paper - related work and introduction
Tweaks in pages 1-6

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{semantic}

@@

\newcommand{\X}{\emph{x}\ }
\newcommand{\Y}{\emph{y}\ }
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}

@@

Many libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario.
Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
way to add a background, or sharing an axis between charts. The alternative is to use a more
low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use ggplot2 \cite{ggplot2}, based on the Grammar of Graphics \cite{grammar}.
A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
properties (\X and \Y coordinate, shape, color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

@@

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

@@

\begin{itemize}
\item Concepts such as bar charts, line charts or charts with aligned
  axes are all expressed in terms of more primitive building blocks using a small number of combinators.
\item The primitives are specified in domain terms. When drawing a line, the value of a \\ \Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
\item Most common chart types can be easily captured as high level abstractions, but there is a
  way of creating a majority of more interesting custom charts.
\item The approach can easily be integrated with the Elm architecture \cite{elm}
	to creating web-based charts that involve animations or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. We outline the structure of an implementation, but omit the details; filling
those in merely requires careful thinking about geometry and projections. An F\#
implementation, including all our examples, is available
at: \urrl{http://github.com/compostjs}.

@@

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$ consisting
of a categorical value $c$ and a number $r$ between $0$ and $1$. As discussed in
Section~\ref{sec:basic-types}, continuous and categorical values can also be annotated with
units of measure to make the values more descriptive.
%
\begin{figure}
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \\
  & | & \kvd{cont}~n\\[1.25em]
  ~
\end{array}
\qquad
\begin{array}{rclcl}
s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
 & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
 & | & \kvd{bubble}~\gamma, v_x, v_y, w, h
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

@@

Now we can define the basic elements of the Compost domain-specific language.
A chart element is represented by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by \X and \Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels rather than
in domain units.

Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
which overlays given shapes. When doing this, Compost infers the range of values on the \X and \Y
axes and calculates suitable projections using a method discussed in the next
section. The $\kvd{padding}$ combinator adds padding around a specified shape and $\kvd{axis}$ adds
an axis showing the inferred scale on the left, right, top or bottom of a given shape.
Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left):
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{conservative},~\ident{labour}~=\\
\quad \kvd{fill}~\strf{\#0000ff},
 ~[\;\,(\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{365}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{365}), (\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{0})\;\,],\\
\quad \kvd{fill}~\strf{\#ff0000},
 ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]\\[0.5em]
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{labour}, \ident{conservative}~]))\\
\end{array}
\end{equation*}

@@

\begin{figure}
  \includegraphics[scale=0.57]{figures/simple}
  \vspace{0.25em}
  \caption{Simple chart showing the UK election results; using automatically inferred scales (left)
    and using rounded Y scale and explicitly defined (reordered) X scale (right).}
  \label{fig:simple}
\end{figure}

@@

\vspace{-0.5em}
\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars. This is specified in pixels rather than in terms of domain values. A similar
result could be be achieved by drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$, but
specifying padding precisely in pixels is sometimes preferable.
The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.

@@

Scales are an important concept in Compost.
When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
to position the child elements relatively to each other. The Compost library positions the elements
automatically. This is done in two steps. During pre-processing, Compost infers the scales for
\X and \Y axes. A scale represents the range of values that needs to fit in the space available for
the chart. When rendering a shape, Compost projects domain-specific values to the available screen
space based on the inferred scale. A scale $l$ is defined in Figure~\ref{fig:scale}.
%
A continuous scale is defined by a minimal and maximal value that need to be mapped to the
available chart space. A categorical scale is defined by a list of individual categorical values.
Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
will use an equal space for each category, regardless of where in this space a shape needs to
appear.

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}
\vspace{-1em}
\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
\label{fig:scale}
\vspace{-1em}
\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
  & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
\end{array}
\end{equation*}
\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
\label{fig:control}
\end{figure}

Scale inference is done by a simple recursive function that walks over the given shape and
constructs two scales for the \X and \Y axis, using the \X and \Y coordinates that appear in the shape.
Most of the work is done by a simple helper function that takes two scales, $l_1$~and $l_2$,
and produces a new scale that represents the union of the two:
%
\begin{equation*}
\begin{array}{l}
\ident{union}~(\kvd{continuous}~n_l, n_h)~(\kvd{continuous}~n'_l, n'_h) =\\
\qquad\kvd{continuous}~\min(n_l, n'_l), \max(n_h, n'_h)\\[0.5em]
\ident{union}~(\kvd{categorical}~[\,c_1, \ldots, c_p\,])~(\kvd{categorical}~[\,c'_1, \ldots, c'_q\,]) =\\
\qquad \kvd{categorical}~[\,c_1, \ldots, c_p\,] \;@\; [\,c'_i~|~ \forall i\in 1\ldots q, \nexists j. c_j = c'_i \;]
\end{array}
\end{equation*}

@@

@@

@@

The automatic scale inference often results in scales where the maximum is a non-round number.
This leads to charts that fully utilise the available space, but may not be easy to read.
The first two primitives, shown in Figure~\ref{fig:control} (left) allow the chart designer
to adjust the automatically inferred range of scales.
The operations can be applied to either the \X scale or the \Y scale, which is indicated by the
$x/y$ subscript. The \kvd{roundScale} primitive takes the inferred \X or \Y scale of the shape $s$
and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
\end{array}
\end{equation*}
\vspace{-0.75em}

@@

The most interesting primitive for controlling scale composition
is $\kvd{nest}_{x/y}$. The behaviour is similar to that of the SVG viewport \cite{svg}.
It takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
constructed from the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continuous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

@@

\begin{figure}
  \includegraphics[scale=0.57]{figures/standard}
  \vspace{0.25em}
  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris\\
    dataset with a title (left) and a pairplot comparing two Iris features (right).}
  \label{fig:standard}
\end{figure}

The second shape simply wraps the specified chart $s$ to which we are attaching the title. The first
positions the text title in the middle of the available space. To do so, we explicitly set the \X and
\Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$.
Figure~\ref{fig:standard} (left) shows a sample scatter plot chart with a title created using the
\ident{title} combinator.

@@

The example shows that the Compost is simple yet expressive. With just
a few lines of code, we are able to construct charts that, in other systems, require dedicated
libraries. The essential aspect of the language making this possible is the automatic inference
of scales and their mapping to the available space as well as the \kvd{nest} operation.

@@

\subsection{Domain-specific events}
Compost can be used in conjunction with a virtual-dom library to create interactive data
visualizations based on the Elm architecture \cite{mvu}. An interactive visualization is described using
a pair of functions -- the \emph{view} function creates a chart based on the current state and
the \emph{update} function specifies how the state changes when an event occurs.

@@

\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{mouseUp}~(\lambda \,x~y \rightarrow e), s  &|& \kvd{mouseDown}~(\lambda \,x~y\rightarrow e), s  \\
    &|& \kvd{mouseMove}~(\lambda \,x~y \rightarrow e), s &|& (\ldots)
\end{array}
\end{equation*}
\caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
\vspace{0.5em}
\label{fig:mouse}
\end{figure}

\subsection{You Draw It data visualizations}
To illustrate building of interactive data visualizations using Compost, we look at one aspect of
``You Draw It''. We want to create a bar chart where the user can use drag \& drop
to move individual bars. Figure~\ref{fig:youdraw} shows the interactive chart before and after an
interaction. The first step is to define types representing the state and events that can occur:

@@

\vspace{-0.5em}
\noindent
The state is a pair of boolean, indicating whether the user is currently dragging, and
a list of key value pairs, storing the number of seats for each political party. Two types of events
can occur in the visualization. First, the user may start or stop dragging, which is indicated using
$\ident{Moving}(\fkvd{true})$ and $\ident{Moving}(\fkvd{false})$, respectively. Second, the user
may change a value for a party, which is represented by the $\ident{Update}$ event.

@@

>>>>

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{semantic}
\usepackage{comment}

\begin{comment}
Dear editors,
I'm grateful to the reviewers for their thorough assessment of the paper. I was very happy to read that the reviewer #2 perfectly captured my hope for the paper ("I must confess that at first Compost seemed too simplistic---but upon reflection I realize that is rather the point"), but the reviewers 1 and 3 made a number of excellent points that deserved clarification. I made a number of changes in response to the reviews.

MAJOR CHANGES
The following two major additions should clarify the paper w.r.t. many of the points made by reviewer 1 about the language design and its expressive power and limitations, as well as to answer points made by other reviewers (Haskell Chart, pie charts):

* I added a new section 2.3 (Choosing the level of language abstraction) which provides more background on the design of the DSL and contrasts it with Haskell Chart library and the Pan language. This should address the concerns of reviewer 1 and explain that, e.g. supporting combinators like "above s1 s2" was not the intention of the design.

* I added a new section 7 (Limitations and further work) that documents some of the things that are currently not supported. This discusses issues such as Pie charts (mentioned by reviewers 1 and 3), logarithmic and contracted axes (reviewer 3) and also visual aspects (reviewer 3). The section explains what is "not yet implemented" vs. what is "not supported by design".

MINOR CHANGES
The following are the notable minor changes that directly address the more specific reviewer comments:

* Reviewer 1 asked for clarification re "nest" - the explanation has been extended to be more clear
* Reviewer 2 pointed out a number of typos - all have been fixed (Thank you!)
* Reviewer 2 asked about figure placement - I tried to improve this (but sometimes unsuccessfuly as I also want to keep long code snippets on one page)
* Reviewer 3 asked about tight links with F# and Elm - I extended the discussion of implementation at the end of Section 1
* Reviewer 3 asked for clarification of syntactic categories in Figure 3 - this is now done.
* Reviewer 3 asked about syntax of host language vs. DSL - this is now explained at the start of Section 2.2
* Reviewer 3 asked about units of measure - I believe this is covered in Section 2.5 (in the context of types)
* Reviewer 3 asked for clarification re interaction - I added new paragraph in Section 5 and also made changes throughout to answer other specific questions raised

\end{comment}

@@

\newcommand{\X}{\emph{x}\ }
\newcommand{\Y}{\emph{y}\ }
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{#1}}}

@@

Many libraries can draw bar charts and line charts, but extra features like multiple bars
for each label, alignment of multiple charts or custom color coding can only be used if the
library author already thought about your exact scenario.
Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
way to add a background or share an axis between charts. The alternative is to use a more
low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use ggplot2 \cite{ggplot2}, based on the Grammar of Graphics \cite{grammar}.
A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
properties (\X and \Y coordinate, shape, color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

\begin{figure}[b]
  \vspace{0.5em}
  \includegraphics[scale=0.55]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about UK politics: Comparison of election results from 2017 and 2019 (left)
  and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

@@

@@

\begin{itemize}
\item Concepts such as bar charts, line charts or charts with aligned
  axes are all expressed in terms of more primitive building blocks using a small number of combinators.
\item The primitives are specified in domain terms. When drawing a line, the value of a \\ \Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
\item Common chart types such as bar charts or line charts can be easily captured as high level
  abstractions, but many interesting custom charts can be created as well.
\item The approach can easily be integrated with the Elm architecture \cite{elm}
	to create web-based charts that involve animations or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. We outline the structure of an implementation, but omit the details; filling
those in merely requires careful thinking about geometry and projections.

Compost is available as open-source at \urrl{http://compostjs.github.io}. It is implemented in F\#,
but is available as a plain JavaScript library thanks to the Fable F\# to JavaScript compiler.
The core logic consists of 800 lines of code and depends on the virtual-dom
library (\urrl{http://npmjs.com/package/virtual-dom}) for the
implementation of the interactive features, meaking it easily portable to other functional
programming languages.

@@

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$ consisting
of a categorical value $c$ and a number $r$ between $0$ and $1$. As discussed in
Section~\ref{sec:basic-types}, continuous and categorical values can also be annotated with
units of measure to make the values more descriptive.
%
\begin{figure}
\begin{equation*}
\begin{array}{llcl}
\textit{Category} & c\\
\textit{Ratio} & r\\
\textit{Number} & n\\
\textit{Text} & t \\
\textit{Color} & \gamma \\
\textit{Value} & v & = & \kvd{cat}~c, r \\
&   & | & \kvd{cont}~n
\end{array}
\begin{array}{lrcl}
\textit{Shape} & s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,]\\
&  & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] \\
&  & | & \kvd{text}~\gamma, v_x, v_y, t \\
&  & | & \kvd{bubble}~\gamma, v_x, v_y, n_w, n_h \\
&  & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,] \\
&  & | &\kvd{axis}_{l/r/t/b}~s\\
&  & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

@@

Compost is an embedded domain-specific language, implemented as a set of functions. In the
subsequent code samples, we will use color to distinguish primitives of the Compost language,
such as \kvd{overlay} or \kvd{cat} from primitives of the host language such as \fkvd{let} or
\fkvd{for}.

A chart element is represented by a shape $s$, as defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by \X and \Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels rather than
in domain units.

Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
which overlays given shapes. When doing this, Compost infers the range of values on the \X and \Y
axes and calculates suitable projections using a method discussed in the next
section. The $\kvd{padding}$ combinator adds padding around a specified shape and $\kvd{axis}$ adds
an axis showing the inferred scale on the left, right, top or bottom of a given shape.
Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left):

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{conservative},~\ident{labour}~=\\
\quad \kvd{fill}~\strf{\#0000ff},
 ~[\;\,(\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{365}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{365}), (\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{0})\;\,],\\
\quad \kvd{fill}~\strf{\#ff0000},
 ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]\\[0.5em]
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{conservative}, \ident{labour}~]))\\
\end{array}
\end{equation*}

@@

\begin{figure}
  \includegraphics[scale=0.55]{figures/simple}
  \vspace{0.25em}
  \caption{Simple chart showing the UK election results; using automatically inferred scales (left)
    and using rounded Y scale and explicitly defined (reordered) X scale (right).}
  \label{fig:simple}
  \vspace{-1.5em}
\end{figure}

@@

\vspace{-0.5em}
\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars, which is specified in pixels. A similar
result could be be achieved by drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$, but
specifying padding precisely in pixels is sometimes preferable.
The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.

\subsection{Choosing the level of language abstraction}
\label{sec:basic-level}

Perhaps the most important aspect of the design of any domain-specific language is the level
of abstraction it uses. The bar chart example discussed in the previous section illustrates
the choice made in Compost. On the one hand, Compost gives us flexibility by letting us
compose charts from shapes. On the other hand, Compost limits what we can do by using
two-dimensional space with positions determined by categorical or continuous values.
In other words, the Compost design lies in the middle of a broader spectrum.

An example of a more general domain-specific language is the Pan language \cite{fun} for producing
images. Pan represents images as functions from a 2D point to a color. This makes it possible to
create powerful combinators, e.g.~polar image transformation, but it makes it harder
to express logic important for charting, e.g.~automatic alignment of shapes defined in terms
of categorical values. Compost makes it easy to put two bars side-by-side in a
bar chart, but harder to define generic combinators for aligning images.

An example of a less general domain-specific language is the Haskell Chart library \cite{hcharts}.
Haskell Chart provides a wide range of plots (such as lines, candles, areas, points, error bars, pies, etc.),
but those can only be composed in limited ways by overlaying them or arranging them in a grid.
The language is closer to the domain of the most common applications, but it places more
restrictions on what can be expressed.

The design of the Compost domain-specific language aims to capture the key principles shared
by most charts, but avoid using a long list of different plot types. Different types of charts
are all produced by composing shapes, but the ways in which shapes can be composed and transformed
are limited to those that are needed for typical charts. We discuss the limitations of this
approach in more detail in Section~\ref{sec:limits}.

@@

Scales are an important concept in Compost.
When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
to position the child elements relatively to each other. The Compost library positions the elements
automatically. This is done in two steps. During pre-processing, Compost infers the scales for
\X and \Y axes. A scale represents the range of values that needs to fit in the space available for
the chart. When rendering a shape, Compost projects domain-specific values to the available screen
space based on the inferred scale. A scale $l$ is defined in Figure~\ref{fig:scale}.
%
A continuous scale is defined by a minimal and maximal value that need to be mapped to the
available chart space. A categorical scale is defined by a list of individual categorical values.
Note that we do not need minimal and maximal ratios of the used categorical values as Compost
will use equal space for each category, regardless of where in this space a shape needs to
appear.

Scale inference is done by a simple recursive function that walks over the given shape and
constructs two scales for the \X and \Y axis, using the \X and \Y coordinates that appear in the shape.
Most of the work is done by a simple helper function that takes two scales, $l_1$~and $l_2$,
and produces a new scale that represents the union of the two:

\begin{equation*}
\begin{array}{l}
\ident{union}~(\kvd{continuous}~n_l, n_h)~(\kvd{continuous}~n'_l, n'_h) =\\
\qquad\kvd{continuous}~\min(n_l, n'_l), \max(n_h, n'_h)\\[0.5em]
\ident{union}~(\kvd{categorical}~[\,c_1, \ldots, c_p\,])~(\kvd{categorical}~[\,c'_1, \ldots, c'_q\,]) =\\
\qquad \kvd{categorical}~[\,c_1, \ldots, c_p\,] \;@\; [\,c'_i~|~ \forall i\in 1\ldots q, \nexists j. c_j = c'_i \;]
\end{array}
\end{equation*}

@@

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{rcl}
\textit{Scale}~~~l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}
\vspace{-1em}
\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
\label{fig:scale}
\vspace{-1em}
\end{figure}

@@

\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
\textit{Shape}~~~s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
   & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
\end{array}
\end{equation*}
\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
\label{fig:control}
\end{figure}

@@

The automatic scale inference often results in scales where the maximum is a non-round number.
This leads to charts that fully utilise the available space, but may not be easy to read.
The first two primitives, shown in Figure~\ref{fig:control} (left) allow the chart designer
to adjust the automatically inferred range of scales.
The operations can be applied to either the \X scale or the \Y scale, which is indicated by the
$x/y$ subscript. The \kvd{roundScale} primitive takes the inferred \X or \Y scale of the shape $s$
and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
following code (reusing the \ident{conservative} and \ident{labour} variables defined earlier):
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
\qquad \kvd{overlay}~[~\ident{conservative}, \ident{labour}~]~)))\\
\end{array}
\end{equation*}
\vspace{-0.75em}

@@

The most interesting primitive for controlling scale composition
is $\kvd{nest}_{x/y}$. As with other primitives like \kvd{padding}, the primitive takes a shape
with some additional parameters and defines a new shape.
Its behaviour is similar to that of the SVG viewport \cite{svg}.
The \kvd{nest} primitive takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~v_{min}, v_{max}, s$ will be a
categorical or continuous scale constructed from the values $v_{min}$ and $v_{max}$, regardless
of the values that are used inside the shape $s$. The chart space between $v_{max}$ and $v_{min}$
will then be used to render the nested shape $s$ using its inferred scale. In other words, the
operation defines a virtual coordinate system that exists only inside the newly created shape,
but is invisible to anything outside of the shape.
An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continuous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

@@

\begin{figure}
  \includegraphics[scale=0.57]{figures/standard}
  \vspace{0.25em}
  \caption{Sample charts built using derived abstractions; a scatter plot visualizing the Iris dataset with a title (left) and a pairplot comparing two Iris features (right).}
  \label{fig:standard}
\end{figure}

The second shape simply wraps the specified chart $s$ to which we are attaching the title. The first
positions the text title in the middle of the available space. To do so, we explicitly set the \X and
\Y scales inside the upper shape to continuous scales from $\num{0}$ to $\num{100}$ and then position
the text label in the middle, at a point $(\kvd{cont}~\num{50}), (\kvd{cont}~\num{50})$. We assume
that the \kvd{text} primitive centers the text, although the actual implementation also allows the
user to specify horizontal and vertical alignment.
Figure~\ref{fig:standard} (left) shows a sample scatter plot chart with a title created using the
\ident{title} combinator.

@@

The example shows that Compost is simple yet expressive. With just
a few lines of code, we are able to construct charts that, in other systems, require dedicated
libraries. The essential aspect of the language making this possible is the automatic inference
of scales and their mapping to the available space as well as the \kvd{nest} operation.

@@

Compost uses the Elm architecture \cite{mvu} to support interactive data visualizations.
In this model, an interactive visualization is described using a pair of user-defined types
and a pair of user-defined functions. The \emph{state} type represents the current state of what
is displayed (e.g.~animation step or selection) and the \emph{event} type represents actions that
the user can perform (e.g.~start an animation or draw a selection range). The two functions
use the state and event types. The \emph{view} function creates a chart based on the current
state and the \emph{update} function specifies how the state changes when an event occurs.

\subsection{Domain-specific events}

@@

\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{mouseUp}~(\lambda \,x~y \rightarrow e), s  &|& \kvd{mouseDown}~(\lambda \,x~y\rightarrow e), s  \\
    &|& \kvd{mouseMove}~(\lambda \,x~y \rightarrow e), s &|& (\ldots)
\end{array}
\end{equation*}
\caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
\vspace{0.5em}
\label{fig:mouse}
\end{figure}

\subsection{You Draw It data visualizations}
To illustrate building interactive data visualizations using Compost, we look at one aspect of
``You Draw It''. We want to create a bar chart where the user can use drag \& drop
to move individual bars. Figure~\ref{fig:youdraw} shows the interactive chart before and after an
interaction. The first step is to define types representing the state and events that can occur:

@@

\vspace{-0.5em}
\noindent
The state is a pair of a boolean, indicating whether the user is currently dragging, and
a list of key/value pairs, storing the number of seats for each political party. Two types of events
can occur in the visualization. First, the user may start or stop dragging, which is indicated using
$\ident{Moving}(\fkvd{true})$ and $\ident{Moving}(\fkvd{false})$, respectively. Second, the user
may change a value for a party, which is represented by the $\ident{Update}$ event.

@@

Finally, the \ident{view} function takes the current state and builds the data visualization
using the Compost domain-specific language. In addition, it also takes a parameter \ident{trigger},
which is an effectful function of type $\ident{Event}\rightarrow\ident{unit}$ that can be used to
trigger events in handlers, registered using primitives such as \kvd{mouseMove}. The \ident{trigger}
function is provided by the Compost runtime. When it is invoked from an event handler, it takes
the current state, transforms it using the \emph{update} function, sets the new state as the
current state and invokes the \emph{view} function to display the new state.
